---
title: "The probabilistic mind: A Bayesian cognitive modeling tutorial"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_download: true
    number_sections: true
    theme: flatly
    highlight: tango
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: inline
---

<font size="4">

# Introduction

## Technical objective
Learn the basics of Bayesian cognitive modeling, a method for simulating the mind's psychological processes as functions of a Bayesian inference machine

## Substantive research question
How do feelings of stress and dominance emerge probabilistically in response to conversations with partners of the same and different gender?

## Emotional reactivity to social interactions
In this tutorial... [TODO]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load greta package for writing and sampling statistical models
library(greta)
# load packagesfor plotting greta models
library(igraph)
library(DiagrammeR)

# install necessary TensorFlow and TensorFlow Probability Python packages
install_greta_deps()
```

```{r person-level data}
# set filepath for data file
filepath <-
  "https://raw.githubusercontent.com/The-Change-Lab/collaborations/main/AMIB/AMIB_persons.csv"
# read in the .csv file using the url() function
AMIB_persons <- read.csv(file=url(filepath),header=TRUE)
```

```{r day-level data}
# set filepath for data file
filepath <-
  "https://raw.githubusercontent.com/The-Change-Lab/collaborations/main/AMIB/AMIB_daily.csv"
# read in the .csv file using the url() function
AMIB_daily <- read.csv(file=url(filepath),header=TRUE)
```

```{r interaction-level data}
# set filepath for data file
filepath <-
  "https://raw.githubusercontent.com/The-Change-Lab/collaborations/main/AMIB/AMIB_interaction.csv"
# read in the .csv file using the url() function
AMIB_interaction <- read.csv(file=url(filepath),header=TRUE)
```

Note: This code doesn't run because WebPPL has not been set up (yet).  

Contributions of personal stress level and partner's gender to dominance in a social interaction:  

TODO: change the partnerGender variable to be [0 = same gender interlocutor vs. 1 = opposite/different gender interlocutor]
TODO: try to write this model in greta (a PPL hosted in R that runs atop TensorFlow inference algos: https://cran.r-project.org/web/packages/greta/vignettes/get_started.html)
TODO: after you get this model up and running in greta, try out a new model with different variables (perhaps some that are more interesting/are supported by prior work - or find prior work supporting the stress/dominance/gender associations)
```{js gender stress dominance}
var getIgdom = function(partnerGender, stress) { // igdom = interaction-level dominance
  if (partnerGender == 1) { // partner is female
    return stress == 1 ? 7 : 5
  } // partner is male
  return stress == 1 ? 6 : 4
}

var model = function () {
    var partnerGender = flip() // equal chance of having male vs. female partner
    var stress = flip(0.3) // 30% chance of being stressed
    var igdom = getIgdom(partnerGender, stress)
    condition(igdom >= 5)
    return partnerGender
}
var dist = Infer({method: 'rejection', samples: 1000}, model)
viz(dist)
```
# TODO: plot the model
# TODO: evaluate the model against the data
# TODO: dig into the data and try to make a model that matches the data (this is an old comment - not really sure what I meant by this)

</font>